---
output:
  pdf_document:
    latex_engine: pdflatex
header-includes:
  - \usepackage{booktabs}
  - \usepackage{tikz}
  - \usepackage{float}
---

```{r setup, include=FALSE}
library(knitr)
raw <- read.csv("asimplied.csv", check.names = FALSE, stringsAsFactors = FALSE)
dep <- format(strptime(raw$SOBT, "%d-%m-%Y %H:%M"), "%H%M")
corpus <- data.frame(flight = raw[["Flight Designator"]],
                     n      = raw[["S.no"]],
                     route  = raw[["Routing"]],
                     dest   = raw[["dest_iata"]],
                     cap    = raw[["Seats"]],
                     dep    = dep,
                     stringsAsFactors = FALSE)
corpus <- corpus[order(corpus$n), ]
```

Scope: International Departures @ Terminal 3 on June 21, 2026

```{r corpus-table, echo=FALSE, results='asis'}
cat("\\begin{table}[H]\n\\centering\n")
cat(kable(head(corpus, 10), format = "latex", booktabs = TRUE,
          row.names = FALSE, align = "llllrl"))
cat("\n\\caption{Excerpt from Corpus, from 0000 hours to 0100 hours}\n\\end{table}\n")
```

The Corpus contains `r nrow(corpus)` flights with total capacity `r format(sum(corpus$cap), big.mark=",")`.

```{r corpus-graph, echo=FALSE, results='asis'}
hr  <- sort(as.integer(substr(corpus$dep, 1, 2)))
k   <- ave(hr, hr, FUN = function(z) seq_along(z) - 1L)
cnt <- as.integer(table(factor(hr, levels = 0:23)))
W <- 15.5; ystep <- 0.18; r <- 1.6; y0 <- 0.22
xof <- function(h) (h + 0.5)/24 * W
cat("\\begin{figure}[H]\n\\centering\n\\begin{tikzpicture}\n")
cat(sprintf("\\draw[gray] (0,0) -- (%.3f,0);\n", W))
for (h in 0:23) {
  x <- xof(h)
  cat(sprintf("\\draw[gray] (%.3f,0) -- (%.3f,-0.10);\n", x, x))
  cat(sprintf("\\node[below] at (%.3f,-0.10) {\\scriptsize %02d};\n", x, h))
}
for (i in seq_along(hr)) {
  cat(sprintf("\\fill (%.3f,%.3f) circle[radius=%.1fpt];\n", xof(hr[i]), y0 + k[i]*ystep, r))
}
for (h in 0:23) if (cnt[h+1] > 0) {
  cat(sprintf("\\node[font=\\scriptsize] at (%.3f,%.3f) {%d};\n", xof(h), y0 + (cnt[h+1]-1)*ystep + 0.26, cnt[h+1]))
}
cat("\\end{tikzpicture}\n\\caption{Distribution of dep(flights) over hourly time, each dot = one flight}\n\\end{figure}\n")
```

We assign a serial number to each occupiable seat, continuous over all flights in order of departure.

```{r seat1-table, echo=FALSE, results='asis'}
seats <- read.csv("seats.csv", check.names = FALSE, stringsAsFactors = FALSE)
sel <- head(seats[seats$seat_on_flight == 1, c("seat", "n", "flight")], 10)
cat("\\begin{table}[H]\n\\centering\n")
cat(kable(sel, format = "latex", booktabs = TRUE, row.names = FALSE, align = "rrl"))
cat("\n\\caption{Excerpt from Seats, first seat for flights n = [1, 10]}\n\\end{table}\n")
```

The actual number of Passengers on a flight is defined by its Load Factor, where

**Equation 1.** Number of passengers: `pax = round(LF * cap)` [for a given flight $n$].

**Assumption 1.** *Homogeneous LF: all flights have an equal load factor that is not idiosyncratic; this will be relaxed later.*

For a given seat, we assign a binary quality `exists = {0, 1}`, where a pax is defined by `exists(seat) = 1`.

A flight's check-in window opens at `(dep - 0400)` and closes at `(dep - 0100)`. Each correspondent pax has a time `entry(pax)`, defined as the time a passenger enters the check-in area.

Check-in desks are 'points' that process pax at a given rate.

**Assumption 2.** *Homogeneous Check-in Rate: all check-in desks have an equal rate of processing that is not idiosyncratic; this will be relaxed later.*

Processing rates are expressed as a service time $\tau$ in seconds per pax, or equivalently a service rate $\mu = 1/\tau$ in pax per unit time.


\begin{center}
\begin{tikzpicture}[x=1cm,y=1cm]
  \draw[line width=1.2pt] (-0.30,0.78) -- (0.20,0.78);
  \draw[line width=1.2pt] (0.40,0.78) -- (0.90,0.78);
  \foreach \lab/\yy in {A/0, B/-0.5, C/-1.0, D/-1.5} {
    \fill (0,\yy) circle (2.5pt);
    \node[left=7pt] at (0,\yy) {\lab};
  }
  \draw[line width=0.8pt] (1.3,-1.5) circle (2.7pt);
  \node[left=7pt] at (1.3,-1.5) {E};
\end{tikzpicture}

\vspace{2mm}
\textbf{Illustration 1.} Basic Setup with 1 desk and 4+1 pax
\end{center}

In the shown arrangement, with a service time of $\tau$ seconds per pax, passenger E would queue for $4\tau$ seconds and be processed within $5\tau$ seconds^[The two bars represent one check-in desk point, where filled dots represent queued passengers and the empty dot represents a passenger yet to queue.].


**Assumption 3.** *Given multiple check-in options, the rational passenger chooses that with the shortest queue and thence the shortest processing time.*

\begin{center}
\begin{tikzpicture}[x=1cm,y=1cm]
  % Desk 1
  \node at (0.25,2.55) {1};
  \draw[line width=1.2pt] (-0.25,2.15) -- (0.20,2.15);
  \draw[line width=1.2pt] (0.30,2.15) -- (0.75,2.15);
  \foreach \lab/\yy in {A/1.5, B/1.0, C/0.5, D/0.0} {
    \fill (0.25,\yy) circle (2.5pt);
    \node[left=7pt] at (0.25,\yy) {\lab};
  }
  % Desk 2
  \node at (3.75,2.55) {2};
  \draw[line width=1.2pt] (3.25,2.15) -- (3.70,2.15);
  \draw[line width=1.2pt] (3.80,2.15) -- (4.25,2.15);
  \foreach \lab/\yy in {X/1.5, Y/1.0, Z/0.5} {
    \fill (3.75,\yy) circle (2.5pt);
    \node[left=7pt] at (3.75,\yy) {\lab};
  }
  % E (unqueued)
  \draw[line width=0.8pt] (2.0,-0.55) circle (2.7pt);
  \node[left=7pt] at (2.0,-0.55) {E};
\end{tikzpicture}

\vspace{2mm}
\textbf{Illustration 2.} Rational Choice: Passenger E chooses to queue for check-in desk 2, ceteris paribus.
\end{center}
