---
title: "asimplied.csv --- Descriptive Statistics"
subtitle: "DEL International Departures, 21 June 2026"
date: "Generated 19 June 2026"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
geometry: margin=1in
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(knitr.kable.NA = "")
library(knitr)

path <- "C:/Users/yashm/OneDrive - Cornell University/Desktop/Work/Delhi International Airport Limited/Pax_Forecast_Model/asimplied.csv"
df <- read.csv(path, stringsAsFactors = TRUE, check.names = FALSE)

clean <- function(s) gsub("_", " ", s)   # tidy underscored field names for display
```

# Overview

This report profiles every column of `asimplied.csv`, the Day-7 international
departures schedule for Delhi (DEL) on 21 June 2026. The dataset contains
**`r nrow(df)` flights** (rows) across **`r ncol(df)` fields** (columns).

```{r overview}
ov <- data.frame(
  Field    = names(df),
  Type     = sapply(df, function(x) if (is.numeric(x)) "numeric" else "categorical"),
  Distinct = sapply(df, function(x) length(unique(x[!is.na(x)]))),
  Missing  = sapply(df, function(x) sum(is.na(x))),
  row.names = NULL
)
kable(ov, format = "latex", row.names = FALSE,
      col.names = c("Field", "Type", "Distinct values", "Missing"))
```

# Numeric Summary: Seats

Seats is the only numeric measure in the schedule. `r sum(!is.na(df$Seats))` of
`r nrow(df)` flights report a seat count (one value is missing).

```{r seats}
s <- df$Seats
seats_tbl <- data.frame(
  Statistic = c("Flights with data", "Missing", "Mean", "Median",
                "Std. deviation", "Minimum", "25th percentile",
                "75th percentile", "Maximum", "Total seats (all flights)"),
  Value = c(
    format(sum(!is.na(s))),
    format(sum(is.na(s))),
    format(round(mean(s, na.rm = TRUE), 1)),
    format(median(s, na.rm = TRUE)),
    format(round(sd(s, na.rm = TRUE), 1)),
    format(min(s, na.rm = TRUE)),
    format(quantile(s, 0.25, na.rm = TRUE)),
    format(quantile(s, 0.75, na.rm = TRUE)),
    format(max(s, na.rm = TRUE)),
    formatC(sum(s, na.rm = TRUE), format = "d", big.mark = ",")
  )
)
kable(seats_tbl, format = "latex", row.names = FALSE,
      col.names = c("Statistic", "Value"))
```

# Categorical Frequencies

For each categorical field the most common values are listed (top 15 where a
field has more than 15 distinct values). Share is the percentage of all 165
flights.

```{r cats, results='asis'}
cat_cols <- c("Airline code", "Traffic Type", "Public Terminal", "GHA",
              "Routing", "dest_iata", "dest_city", "dest_country",
              "airline_name", "airline_country")
for (col in cat_cols) {
  tb    <- sort(table(df[[col]]), decreasing = TRUE)
  n_lvl <- length(tb)
  top   <- head(tb, 15)
  d <- data.frame(
    Value = names(top),
    Count = as.integer(top),
    Share = paste0(formatC(100 * as.integer(top) / sum(tb),
                           format = "f", digits = 1), "%")
  )
  cat("\n\n## ", clean(col), " (", n_lvl, " distinct)\n\n", sep = "")
  tab <- kable(d, format = "latex", row.names = FALSE,
               col.names = c(clean(col), "Count", "Share"))
  cat(tab, sep = "\n")
  if (n_lvl > 15) cat("\n\n*Top 15 of ", n_lvl, " levels shown.*", sep = "")
  cat("\n\n")
}
```
