---
title: "DIAL Terminal 3 International Departures"
subtitle: "A Dynamic Stocks and Flows Model of Passenger Arrival, Check-in Queueing, and Counter Assignment"
author: "Prepared for Delhi International Airport Limited"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: kable
    latex_engine: pdflatex
geometry: margin=1in
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{fancyhdr}
  - \usepackage{lastpage}
  - \usepackage{booktabs}
  - \pagestyle{fancy}
  - \fancyhf{}
  - \rhead{DIAL T3 Check-in Model}
  - \lhead{Stocks and Flows}
  - \cfoot{Page \thepage\ of \pageref{LastPage}}
  - \renewcommand{\footrulewidth}{0.4pt}
params:
  dt: 1
  load_factor: 0.90
  seed: 42
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 4
)
options(stringsAsFactors = FALSE)
set.seed(params$seed)
```

# Scope and reading guide

This document specifies and implements a single, fully parameterised model of the
passenger check-in process for international departures from Terminal 3 (T3) at
Delhi airport. Everything in scope here is T3 international only. The model is
written as a stocks and flows system.^[A stocks and flows system tracks
quantities that accumulate over time (stocks, for example the number of people
currently waiting in a queue) and the rates that change them (flows, for example
people joining the queue and people being processed). It is the standard way to
describe a queue that fills and drains over time.]

Every assumption is isolated, named, and changeable. The same engine runs a
zero-assumption ideal case and progressively richer cases by swapping one
configuration object. The text uses mathematical notation for precision;
footnotes translate technical terms into plain language where useful.

The order of the document is: glossary, entity definitions, notation, governing
equations, the configurable engine, solved cases, verification, and a closing
agenda of items to confirm with leadership.

\newpage

# Glossary

The following terms are used with fixed meanings throughout.

| Term | Meaning |
|------|---------|
| SOBT | Scheduled Off-Block Time. The scheduled departure datum, written $T$. |
| Check-in window | The interval during which a flight accepts check-in, fixed at $[T-4\text{h},\,T-1\text{h}]$. |
| Open / Close | Check-in opens at $T-4$h and closes at $T-1$h. These are fixed. |
| Show-up | A passenger who actually arrives to check in for the flight. |
| Entry | The instant a passenger joins the check-in queue, written $e_i$. |
| Service begin | The instant a counter starts processing a passenger, written $b_i$. |
| Processed | The instant a passenger finishes check-in, written $d_i$. |
| Lag | Time from entry to processed, $\ell_i = d_i - e_i$. |
| Wait | Time from entry to service begin, $w_i = b_i - e_i$. |
| No-show | A passenger who arrives after Close. Only possible once the after-Close tail is allowed. |
| Left-behind | A passenger who entered before Close but was still queued at Close. |
| GHA | Ground Handling Agent. The firm staffing the counters (for example AISATS, BFS). |
| Counter / Desk | One check-in position. One counter equals one staff member. |
| Lane | A class restriction on a counter, for example a business-only desk. |
| Load factor | Fraction of seats that become show-ups, written $\rho$ (a variable assumption). |
| Class | A service segment a passenger belongs to, written $\kappa_i$ (economy, premium, business, first, fast-track, loyalty). |
| $\Delta t$ | The model time step in minutes. Variable, default `r params$dt`. |
| Idiosyncrasy | A rule that makes one group differ from the base case (by airline, destination, or departure time). |

\newpage

# Entities and their attributes

We define four entities: the passenger, the flight, the counter, and the service
group. Each is a row in a table whose columns are the attributes below. All
symbols are reused consistently for the rest of the document.

## The passenger $i$

A passenger $i$ carries the following attributes.

| Symbol | Attribute | Meaning |
|--------|-----------|---------|
| $i$ | id | Unique passenger identifier. |
| $f(i)$ | flight | The flight this passenger is booked on. |
| $a(i)$ | airline | Airline of the flight, $a(i)=a_{f(i)}$. |
| $r(i)$ | route | Destination routing of the flight. |
| $g(i)$ | GHA | Ground handler of the flight, $g(i)=g_{f(i)}$. |
| $\kappa_i$ | class | Service class or lane membership (see Section 8). |
| $L_i$ | loyalty | Indicator, $1$ if loyalty-priority, else $0$. |
| $X_i$ | fast-track | Indicator, $1$ if fast-track product held, else $0$. |
| $e_i$ | entry time | Minute at which the passenger joins the queue. |
| $b_i$ | service begin | Minute a counter starts processing the passenger. |
| $d_i$ | processed time | Minute the passenger finishes check-in. |
| $w_i$ | wait | $w_i = b_i - e_i$. |
| $\ell_i$ | lag | $\ell_i = d_i - e_i$. |
| $m_i$ | margin | Buffer minutes the passenger targets before Close (Section 9). |
| $s_i$ | status | One of served, no-show, left-behind. |

## The flight $f$

| Symbol | Attribute | Meaning |
|--------|-----------|---------|
| $\varphi_f$ | designator | Flight number, for example AI 173. |
| $a_f$ | airline | Operating carrier code. |
| $r_f$ | route | Destination routing. |
| $S_f$ | SOBT | Scheduled off-block time in minutes from a reference midnight. |
| $H_f$ | hour | Clock hour of $S_f$. |
| $K_f$ | seats | Physical seat capacity. |
| $\rho$ | load factor | Show-up rate, variable assumption. |
| $N_f$ | show-ups | $N_f = \mathrm{round}(\rho\,K_f)$. Area under the arrival curve. |
| $g_f$ | GHA | Assigned ground handler. |
| $o_f$ | open | $o_f = S_f - 240$ minutes. |
| $c_f$ | close | $c_f = S_f - 60$ minutes. |
| $\mathcal{D}_f$ | arrival law | Arrival distribution family and parameters (Section 7). |
| $n_{f,\kappa}$ | class split | Show-ups in class $\kappa$, with $\sum_\kappa n_{f,\kappa}=N_f$. |

## The counter $j$

| Symbol | Attribute | Meaning |
|--------|-----------|---------|
| $j$ | id | Unique counter identifier. |
| $g_j$ | GHA | Ground handler operating the counter. |
| $a_j$ | airline | Airline the counter is assigned to serve. |
| $\mu_j$ | rate | Service rate in passengers per minute. Service time per passenger is $1/\mu_j$. |
| $o_j$ | open | Minute the counter is staffed from. |
| $c_j$ | close | Minute the counter stops, bounded by the served flights' Close. |
| $\mathcal{A}_j$ | lane | Set of classes the counter may serve. |

## The service group

Counters are pooled by ground handler and then by airline, not by flight.^[This
matches operations: a handler staffs a bank of desks for an airline, and any of
that airline's departing passengers in the window can use any of those desks.]
A service group is the key $(g,a)$. Every passenger with $g(i)=g$ and $a(i)=a$
shares the counter pool of group $(g,a)$. Setting the group key to $(g,a,f)$
recovers a per-flight model, which the engine also supports.

\newpage

# Constraints and conventions

The model enforces the following.

1. Fixed window. Check-in opens at $o_f = S_f-240$ and closes at $c_f = S_f-60$
   for every flight. These are not adjustable.

2. Mass conservation. The area under each flight's arrival curve equals its
   show-ups:
   $$ \int_{o_f}^{c_f} \lambda_f(t)\, dt \;=\; N_f, \qquad N_f=\mathrm{round}(\rho K_f), $$
   where $\lambda_f(t)$ is the arrival intensity (passengers per minute) for
   flight $f$.^[Intensity is how many passengers arrive per minute at time $t$.
   Its total area across the window is the number of people who show up.] When
   the after-Close tail is allowed, the integral runs beyond $c_f$ and the excess
   mass becomes no-shows.

3. Per-passenger deadline. A passenger can only be processed up to his own
   flight's Close. If $b_i \ge c_{f(i)}$ the passenger is left-behind.

4. Resource budget. Counters are scarce. The total counter-minutes consumed,
   $\sum_j (c_j - o_j)$, must not exceed an operating budget $B$. Adding counters
   shortens the time each can be assigned elsewhere.

5. Ideal feasibility. The ideal constraint is zero left-behind and zero no-show.
   A scenario that violates it is feasible to run but flagged.

\newpage

# Governing equations (stocks and flows)

Fix a service group and a class lane. Let $Q(t)$ be the queue stock, the number
of passengers who have entered but not yet begun service at time $t$. Let
$\alpha(t)$ be the arrival flow into the queue and $\delta(t)$ the departure flow
out of the queue into service. The stock evolves as
$$ Q(t+\Delta t) \;=\; Q(t) \;+\; \alpha(t)\,\Delta t \;-\; \delta(t)\,\Delta t, \qquad Q(t)\ge 0. $$

The departure flow is capped by the available service capacity of the open
counters serving that lane:
$$ \delta(t)\,\Delta t \;=\; \min\!\Big( Q(t) + \alpha(t)\Delta t,\; \textstyle\sum_{j \in \text{open}(t)} \mu_j \,\Delta t \Big). $$

The arrival flow is the discretised arrival law,
$\alpha(t)\,\Delta t = N_f \cdot \int_{t}^{t+\Delta t} \hat\lambda_f(u)\,du$, where
$\hat\lambda_f$ is the normalised arrival density. Conservation across the whole
window returns $\sum_t \alpha(t)\Delta t = N_f$ by construction.

The per-passenger lag is read off the service schedule:
$\ell_i = d_i - e_i$ and $w_i = b_i - e_i$. The engine resolves the order of
service inside each step so that these are defined for every individual, which is
the quantity the planner ultimately cares about.

\newpage

# The engine: time, schedule, and arrival laws

```{r time-grid}
# ---------------------------------------------------------------------------
# Time grid with variable resolution. dt divides 60. The window is 180 minutes.
# dissolve() refines a coarse count vector to a finer step (spread equally).
# amalgamate() aggregates a fine count vector to a coarser step (sum).
# These let a distribution defined at one resolution be moved to another.
# ---------------------------------------------------------------------------
WINDOW_MIN <- 180L          # T-4 to T-1 is 180 minutes. Fixed.

make_grid <- function(open_min, dt) {
  M <- WINDOW_MIN / dt
  if (M != floor(M)) stop("dt must divide 180")
  list(dt = dt, M = as.integer(M),
       edges  = open_min + (0:M) * dt,
       centers = open_min + ((1:M) - 0.5) * dt)
}

dissolve <- function(counts, factor) {
  # split each coarse bin into 'factor' finer bins, mass spread equally
  as.numeric(rep(counts / factor, each = factor))
}
amalgamate <- function(counts, factor) {
  # merge every 'factor' fine bins into one coarse bin
  as.numeric(tapply(counts, rep(seq_len(length(counts) / factor), each = factor), sum))
}

# Largest-remainder rounding: round a real vector to integers summing to N.
round_to_total <- function(x, N) {
  if (sum(x) <= 0) { out <- rep(0L, length(x)); if (N > 0) out[1] <- N; return(out) }
  x <- x / sum(x) * N
  fl <- floor(x); rem <- N - sum(fl)
  if (rem > 0) {
    o <- order(x - fl, decreasing = TRUE)[seq_len(rem)]
    fl[o] <- fl[o] + 1L
  }
  as.integer(fl)
}
```

```{r schedule}
# ---------------------------------------------------------------------------
# T3 international departure schedule (the 24 flights). SOBT in minutes from
# midnight. GHA is taken as given here; the per-airline lookup from the data
# build is reused. seats = K_f.
# ---------------------------------------------------------------------------
raw <- "desig|sobt|route|seats|gha
AI 173|0:05|DEL-CCU-SFO|288|AISATS
AI 312|0:05|DEL-ICN|259|AISATS
VJ 972|0:05|DEL-HAN|377|BFS
AI 187|0:10|DEL-VIE-YYZ|328|AISATS
AI 161|0:30|DEL-LHR|316|AISATS
AI 2336|0:30|DEL-BKK|164|AISATS
AF 283|0:40|DEL-SIN|188|AISATS
AI 302|0:55|DEL-SYD|328|BFS
AI 277|1:00|DEL-CMB|259|AISATS
LX 147|1:05|DEL-ZRH|182|AISATS
CX 694|1:10|DEL-HKG|236|BFS
KC 964|1:10|DEL-ALA|280|AISATS
AI 308|1:15|DEL-MEL|179|AISATS
AI 2390|1:20|DEL-HAN|164|AISATS
AI 2378|1:25|DEL-HKT|164|AISATS
BA 142|1:35|DEL-LHR|272|Ramp 360
LX 2647|1:40|DEL-ZRH|235|BFS
LH 763|1:45|DEL-MUC|509|BFS
AI 4307|2:00|DEL-DXB|259|AISATS
VS 303|2:05|DEL-LHR|397|BFS
ET 687|2:30|DEL-ADD|315|Ramp 360
6E 033|2:40|DEL-MAN|318|IndiGo
LH 761|2:50|DEL-FRA|287|BFS
AI 2147|2:55|DEL-DPS|188|AISATS"

flights <- read.table(text = raw, header = TRUE, sep = "|",
                      quote = "", strip.white = TRUE)
hm_to_min <- function(s) {
  p <- do.call(rbind, strsplit(s, ":", fixed = TRUE))
  as.integer(p[, 1]) * 60L + as.integer(p[, 2])
}
flights$airline <- sub("^([0-9A-Z]{2,3})\\s.*$", "\\1", flights$desig)
flights$S  <- hm_to_min(flights$sobt)
flights$H  <- flights$S %/% 60L
flights$o  <- flights$S - 240L     # open  = T-4h
flights$c  <- flights$S - 60L      # close = T-1h
flights$fid <- seq_len(nrow(flights))
```

```{r arrival-laws}
# ---------------------------------------------------------------------------
# Arrival-law registry. Each function returns an UNNORMALISED density evaluated
# at the bin centres x (minutes since open), on [0, W]. They are normalised and
# scaled to N show-ups by arrival_counts().
# Families: uniform, normal, linear, skew (complex), data (the given 35/40/25).
# ---------------------------------------------------------------------------
arrival_density <- list(
  uniform = function(x, W, p) rep(1, length(x)),
  normal  = function(x, W, p) {
    mu <- p$mean %||% (W / 2); sd <- p$sd %||% (W / 4)
    dnorm(x, mu, sd)
  },
  linear  = function(x, W, p) {
    a <- p$intercept %||% 1; b <- p$gradient %||% 0
    pmax(a + b * x, 0)
  },
  skew    = function(x, W, p) {
    xi <- p$xi %||% (W / 3); om <- p$omega %||% (W / 4); al <- p$alpha %||% 4
    z <- (x - xi) / om
    2 / om * dnorm(z) * pnorm(al * z)
  },
  data    = function(x, W, p) {
    # piecewise-constant over thirds of the window: T-4..T-3, T-3..T-2, T-2..T-1
    mass <- p$mass %||% c(0.35, 0.40, 0.25)
    third <- pmin(floor(x / (W / 3)) + 1, 3)
    mass[third]
  }
)
`%||%` <- function(a, b) if (is.null(a)) b else a

arrival_counts <- function(spec, grid, N) {
  W <- grid$M * grid$dt
  fam <- arrival_density[[spec$family]]
  if (is.null(fam)) stop(sprintf("unknown arrival family '%s'", spec$family))
  x_local <- grid$centers - grid$edges[1]      # minutes since open, 0 to W
  dens <- fam(x_local, W, spec$params %||% list())
  if (all(dens <= 0)) dens <- rep(1, length(dens))
  round_to_total(dens, N)         # integer arrivals per bin, summing to N
}
```

The arrival-law registry is the first swappable block. `uniform` is the base
case. `normal` takes a mean and standard deviation. `linear` takes an intercept
and a gradient. `skew` is the complex family, a skew-normal with location, scale,
and shape. `data` reproduces the 35/40/25 split given in the source workbook,
which is retained only as a reference case and is, as a model of true arrivals, a
weak assumption.^[The source split assigns fixed shares to three one-hour blocks
and says nothing about within-hour timing, so it cannot represent build-up or
rush behaviour. It is included for comparison, not as a recommended law.]

# The engine: classes, counters, and policies

```{r class-seg}
# ---------------------------------------------------------------------------
# Class segmentation registry. Each entry gives class fractions (summing to 1)
# and a priority order (higher served first). uni = one lane; bi/tri/quad/quint
# add segments. Fractions are defaults and are overrideable per scenario.
# ---------------------------------------------------------------------------
class_seg <- list(
  uni   = list(frac = c(all = 1.00),
               prio = c(all = 1)),
  bi    = list(frac = c(biz = 0.15, eco = 0.85),
               prio = c(biz = 2, eco = 1)),
  bi_ff = list(frac = c(fast = 0.20, normal = 0.80),
               prio = c(fast = 2, normal = 1)),
  tri   = list(frac = c(biz = 0.10, prem = 0.15, eco = 0.75),
               prio = c(biz = 3, prem = 2, eco = 1)),
  quad  = list(frac = c(first = 0.03, biz = 0.10, prem = 0.15, eco = 0.72),
               prio = c(first = 4, biz = 3, prem = 2, eco = 1)),
  quint = list(frac = c(first = 0.03, biz = 0.10, prem = 0.12, eco = 0.70, loyalty = 0.05),
               prio = c(first = 5, biz = 4, prem = 3, eco = 1, loyalty = 2))
)
```

```{r counters}
# ---------------------------------------------------------------------------
# Counter construction for a service group.
#   n        number of counters
#   mu       service rate in pax per minute (1/mu = minutes per pax)
#   open_pol "all_T4" | "planned" | "responsive"
#   lane_pol "shared" | "dedicated"
# Opening policies set o_j; lane policies set the allowed class set per counter.
# group_open, group_close are the earliest open and latest close in the group.
# ---------------------------------------------------------------------------
build_counters <- function(n, mu, group_open, group_close,
                           open_pol = "all_T4", planned_steps = NULL,
                           lane_pol = "shared", classes = "all", prio = c(all = 1),
                           class_counts = NULL, arrivals_cum = NULL) {
  n <- max(1L, as.integer(n))
  # ---- opening times o_j ----
  if (open_pol == "all_T4") {
    o <- rep(group_open, n)
  } else if (open_pol == "planned") {
    steps <- planned_steps %||% c(0, 30, 60)          # minutes after group_open
    o <- group_open + steps[((seq_len(n) - 1) %% length(steps)) + 1]
  } else if (open_pol == "responsive") {
    # open counter k when cumulative arrivals first reach k * trigger
    trig <- (sum(class_counts %||% 0) %||% 0)
    if (!is.null(arrivals_cum) && length(arrivals_cum) > 0) {
      per <- max(arrivals_cum) / n
      thr <- per * (seq_len(n) - 1)
      idx <- sapply(thr, function(t) which(arrivals_cum >= t)[1])
      o <- group_open + (idx - 1)                      # one bin per minute baseline
    } else o <- rep(group_open, n)
  } else stop("unknown open policy")
  o <- pmin(pmax(o, group_open), group_close - 1)
  # ---- lanes A_j ----
  if (lane_pol == "shared" || identical(classes, "all")) {
    allowed <- replicate(n, classes, simplify = FALSE)
  } else if (lane_pol == "dedicated") {
    cc <- class_counts %||% setNames(rep(1, length(classes)), classes)
    share <- round_to_total(as.numeric(cc[classes]), n)
    share[share == 0] <- 0
    # guarantee at least one counter to any present class if room allows
    allowed <- vector("list", n); k <- 1
    for (cl in classes) {
      reps <- share[match(cl, classes)]
      if (reps > 0) for (r in seq_len(reps)) { allowed[[k]] <- cl; k <- k + 1 }
    }
    while (k <= n) { allowed[[k]] <- classes; k <- k + 1 }   # spare desks are shared
  } else stop("unknown lane policy")
  data.frame(j = seq_len(n), o = o, close = group_close,
             mu = mu, stringsAsFactors = FALSE) -> cdf
  attr(cdf, "allowed") <- allowed
  attr(cdf, "prio")    <- prio
  cdf
}
```

```{r simulator}
# ---------------------------------------------------------------------------
# Event-driven multi-server priority simulator for one service group.
# pax: data.frame(id, e, class, prio, deadline). counters from build_counters().
# Returns pax with b, d, wait, lag, status. Whenever a counter is free it takes
# the highest-priority eligible waiting passenger (then earliest entry).
# ---------------------------------------------------------------------------
simulate_group <- function(pax, counters, eps = 1e-9) {
  n <- nrow(pax)
  if (n == 0) {
    pax$b <- numeric(0); pax$d <- numeric(0)
    pax$wait <- numeric(0); pax$lag <- numeric(0); pax$status <- character(0)
    return(pax)
  }
  allowed <- attr(counters, "allowed")
  cls_levels   <- unique(pax$class)
  pax_code     <- match(pax$class, cls_levels)
  allowed_code <- lapply(allowed, function(a) match(a, cls_levels))
  pax$b <- NA_real_; pax$d <- NA_real_; pax$status <- "served"
  in_wait <- rep(FALSE, n); done <- rep(FALSE, n)
  avail <- counters$o                       # next-free time per counter
  arr_order <- order(pax$e); next_arr <- 1L
  now <- min(c(counters$o, pax$e))
  maxiter <- 50L * (n + nrow(counters)) + 200L; it <- 0L
  repeat {
    it <- it + 1L; if (it > maxiter) break
    # admit arrivals up to now; classify after-close arrivals as no-shows
    while (next_arr <= n && pax$e[arr_order[next_arr]] <= now + eps) {
      idx <- arr_order[next_arr]
      if (pax$e[idx] > pax$deadline[idx] + eps) { pax$status[idx] <- "noshow"; done[idx] <- TRUE }
      else in_wait[idx] <- TRUE
      next_arr <- next_arr + 1L
    }
    # left-behind: waiting and own deadline reached
    lb <- which(in_wait & pax$deadline <= now + eps)
    if (length(lb)) { pax$status[lb] <- "leftbehind"; in_wait[lb] <- FALSE; done[lb] <- TRUE }
    # assign free, open, not-closed counters
    assigned <- FALSE
    free_ctr <- which(avail <= now + eps & counters$o <= now + eps & counters$close > now + eps)
    for (j in free_ctr) {
      cand <- which(in_wait & pax$e <= now + eps & pax$deadline > now + eps &
                    (pax_code %in% allowed_code[[j]]))
      if (!length(cand)) next
      ord <- cand[order(-pax$prio[cand], pax$e[cand])]
      placed <- FALSE
      for (i in ord) {
        b <- max(now, pax$e[i], counters$o[j])
        if (b + eps >= pax$deadline[i] || b + eps >= counters$close[j]) {
          pax$status[i] <- "leftbehind"; in_wait[i] <- FALSE; done[i] <- TRUE
          next
        }
        d <- b + 1 / counters$mu[j]
        pax$b[i] <- b; pax$d[i] <- d; avail[j] <- d
        in_wait[i] <- FALSE; done[i] <- TRUE; assigned <- TRUE; placed <- TRUE
        break
      }
    }
    if (all(done)) break
    if (!assigned) {
      tt <- c()
      if (next_arr <= n) tt <- c(tt, pax$e[arr_order[next_arr]])
      bz <- avail[avail > now + eps & counters$close > now + eps]; if (length(bz)) tt <- c(tt, min(bz))
      op <- counters$o[counters$o > now + eps];                    if (length(op)) tt <- c(tt, min(op))
      wd <- pax$deadline[in_wait];                                 if (length(wd)) tt <- c(tt, min(wd))
      if (!length(tt)) break
      nt <- min(tt); now <- if (nt <= now + eps) now + 1e-6 else nt
    }
  }
  rem <- which(in_wait); if (length(rem)) pax$status[rem] <- "leftbehind"
  pax$wait <- pax$b - pax$e
  pax$lag  <- pax$d - pax$e
  pax
}
```

The simulator is exact for the priority and lane rules: it advances only to
instants where something can change (an arrival, a counter freeing, a counter
opening, or a deadline), and at each such instant gives every free desk the
best eligible passenger it is allowed to serve. The queue stock $Q(t)$ is then
read back from entry and service-begin times for any time grid the planner
chooses.

# The engine: scenario assembly

```{r scenario-runner}
# ---------------------------------------------------------------------------
# A scenario is one list holding every assumption. run_scenario() expands the
# schedule into passengers, groups them, builds counters, simulates, and
# returns the passenger-level table plus per-group counter metadata.
# ---------------------------------------------------------------------------
default_scenario <- function() list(
  dt          = params$dt,
  load_factor = params$load_factor,
  arrival     = list(family = "uniform", params = list()),
  segment     = "uni",
  lane_pol    = "shared",
  open_pol    = "all_T4",
  planned_steps = c(0, 30, 60),
  mu          = 0.40,                 # 2.5 minutes per passenger
  counters_per_group = 6L,
  group_key   = c("gha", "airline"),  # segment by GHA then airline, not flight
  arrival_by  = NULL                  # optional idiosyncrasy overrides (Section 11)
)

expand_passengers <- function(flights, sc) {
  rows <- list(); pid <- 0L
  seg <- class_seg[[sc$segment]]
  for (q in seq_len(nrow(flights))) {
    fl <- flights[q, ]
    N  <- as.integer(round(sc$load_factor * fl$seats))
    grid <- make_grid(fl$o, sc$dt)
    # idiosyncratic arrival law (Section 11), else the scenario default
    spec <- resolve_arrival(sc, fl)
    abin <- arrival_counts(spec, grid, N)
    # class counts for this flight
    classes <- names(seg$frac)
    ccount  <- round_to_total(as.numeric(seg$frac), N); names(ccount) <- classes
    class_pool <- rep(classes, times = ccount)
    if (length(class_pool) < N) class_pool <- c(class_pool, rep(classes[1], N - length(class_pool)))
    class_pool <- class_pool[seq_len(N)]
    # entry times: spread each bin's arrivals evenly inside the bin
    e <- numeric(0)
    for (k in seq_len(grid$M)) {
      mk <- abin[k]; if (mk > 0)
        e <- c(e, grid$edges[k] + (seq_len(mk) - 0.5) / mk * grid$dt)
    }
    if (length(e) < N) e <- c(e, rep(fl$c - 1, N - length(e)))
    e <- sort(e[seq_len(N)])
    cls <- class_pool
    rows[[length(rows) + 1]] <- data.frame(
      id = pid + seq_len(N), fid = fl$fid, airline = fl$airline, gha = fl$gha,
      route = fl$route, H = fl$H, e = e, class = cls,
      prio = unname(seg$prio[cls]), deadline = fl$c, stringsAsFactors = FALSE)
    pid <- pid + N
  }
  do.call(rbind, rows)
}

run_scenario <- function(flights, sc) {
  pax <- expand_passengers(flights, sc)
  key <- interaction(pax[, sc$group_key, drop = FALSE], drop = TRUE)
  seg <- class_seg[[sc$segment]]
  out <- list(); meta <- list()
  for (lv in levels(key)) {
    sub <- pax[key == lv, , drop = FALSE]
    go <- min(sub$e); gc <- max(sub$deadline)
    cc <- table(factor(sub$class, levels = names(seg$frac)))
    arr_cum <- cumsum(tabulate(pmax(1, ceiling(sub$e - go + 1)), nbins = ceiling(gc - go) + 1))
    ctr <- build_counters(
      n = sc$counters_per_group, mu = sc$mu, group_open = go, group_close = gc,
      open_pol = sc$open_pol, planned_steps = sc$planned_steps,
      lane_pol = sc$lane_pol, classes = names(seg$frac), prio = seg$prio,
      class_counts = cc, arrivals_cum = arr_cum)
    res <- simulate_group(sub, ctr)
    res$group <- lv
    out[[lv]] <- res
    meta[[lv]] <- data.frame(group = lv, n_counters = nrow(ctr),
                             counter_minutes = sum(ctr$close - ctr$o))
  }
  list(pax = do.call(rbind, out), counters = do.call(rbind, meta), scenario = sc)
}
```

```{r idiosyncrasy}
# ---------------------------------------------------------------------------
# Idiosyncrasy layer. resolve_arrival() lets the arrival law vary by airline,
# route (destination), or departure hour. arrival_by is a list of override
# rules; the first match wins, else the scenario default applies. This is also
# where geographic concentration in an isodistance radius would shift the law
# (for example a tighter, later peak for a near catchment).
# ---------------------------------------------------------------------------
resolve_arrival <- function(sc, fl) {
  if (is.null(sc$arrival_by)) return(sc$arrival)
  for (rule in sc$arrival_by) {
    ok <- TRUE
    if (!is.null(rule$airline) && !(fl$airline %in% rule$airline)) ok <- FALSE
    if (!is.null(rule$route)   && !(fl$route   %in% rule$route))   ok <- FALSE
    if (!is.null(rule$hour)    && !(fl$H       %in% rule$hour))    ok <- FALSE
    if (ok) return(rule$arrival)
  }
  sc$arrival
}
```

```{r kpis}
# ---------------------------------------------------------------------------
# KPI summary and conservation checks for a scenario result.
# ---------------------------------------------------------------------------
kpis <- function(res) {
  p <- res$pax
  served <- p$status == "served"
  data.frame(
    passengers   = nrow(p),
    served       = sum(served),
    left_behind  = sum(p$status == "leftbehind"),
    no_show      = sum(p$status == "noshow"),
    mean_wait    = round(mean(p$wait[served], na.rm = TRUE), 2),
    p90_wait     = round(as.numeric(quantile(p$wait[served], 0.90, na.rm = TRUE)), 2),
    max_wait     = round(max(p$wait[served], na.rm = TRUE), 2),
    mean_lag     = round(mean(p$lag[served], na.rm = TRUE), 2),
    counter_min  = sum(res$counters$counter_minutes)
  )
}

queue_series <- function(res, grid_dt = 5L) {
  p <- res$pax
  t0 <- floor(min(p$e)); t1 <- ceiling(max(p$deadline))
  ts <- seq(t0, t1, by = grid_dt)
  exit <- ifelse(p$status == "served", p$b, p$deadline)
  Q <- sapply(ts, function(t) sum(p$e <= t & exit > t))
  data.frame(t = ts, Q = Q)
}
```

\newpage

# Lanes, classes, and the home-departure margin

The class segmentation in Section 8 doubles as the lane structure. Under the
`shared` lane policy every counter serves every class and the priority order
alone decides who is taken first. Under the `dedicated` policy counters are
split across classes in proportion to class load, with any spare desks left
shared.^[Dedicated lanes can lower the wait for premium passengers while raising
it for economy, because capacity is reserved. The model reports both so the
trade is visible.]

Class is also the natural carrier of the home-departure margin. A passenger
targets an arrival time and leaves home earlier by a travel time drawn from an
isodistance or isochrone radius around the airport, adjusted for traffic and
time of day.^[An isodistance radius groups origins by straight-line distance; an
isochrone groups them by travel time, which is the more useful notion in
traffic. A traffic and time adjusted isochrone stretches that travel time by the
congestion expected at the relevant hour.] Formally, for passenger $i$ with
target arrival $t^{*}_i$ and travel time $\tau_i$,
$$ h_i \;=\; t^{*}_i - \tau_i, \qquad e_i \;=\; t^{*}_i + \varepsilon_i, $$
where $h_i$ is the leave-home time and $\varepsilon_i$ is small arrival noise.
Higher classes tend to choose a larger margin, so $t^{*}_i$ shifts earlier for
them. The current build draws $e_i$ directly from the arrival law; the margin
path above is the hook through which the isochrone work will feed $e_i$ once the
catchment data is available, without changing anything downstream.

\newpage

# Solved cases

We run two cases through the identical engine and compare. Case S0 is the
zero-assumption ideal: uniform arrivals, one lane, all desks open at $T-4$, equal
service speed, six desks per group. Case S-data swaps in the workbook's 35/40/25
arrival split and nothing else, isolating the effect of that one assumption.

```{r run-cases}
S0 <- default_scenario()

S_data <- default_scenario()
S_data$arrival <- list(family = "data", params = list(mass = c(0.35, 0.40, 0.25)))

res0   <- run_scenario(flights, S0)
resd   <- run_scenario(flights, S_data)

k0 <- kpis(res0); k0$case <- "S0 uniform (ideal target)"
kd <- kpis(resd); kd$case <- "S-data 35/40/25 split"
comp <- rbind(k0, kd)[, c("case","passengers","served","left_behind","no_show",
                          "mean_wait","p90_wait","max_wait","mean_lag","counter_min")]
knitr::kable(comp, caption = "Case comparison through one shared engine.")
```

```{r plot-arrivals, fig.cap="Arrival intensity per 5 minutes: uniform base versus the workbook split, summed across all flights on a common pre-departure axis."}
arr_profile <- function(sc, bin = 5L) {
  acc <- numeric(WINDOW_MIN / bin)
  for (q in seq_len(nrow(flights))) {
    fl <- flights[q, ]
    N <- as.integer(round(sc$load_factor * fl$seats))
    g <- make_grid(fl$o, bin)
    acc <- acc + arrival_counts(resolve_arrival(sc, fl), g, N)
  }
  acc
}
xb <- seq(240, 60, length.out = WINDOW_MIN / 5)   # minutes before departure
a0 <- arr_profile(S0); ad <- arr_profile(S_data)
plot(xb, a0, type = "l", lwd = 2, xlim = c(240, 60),
     xlab = "Minutes before departure (T minus)", ylab = "Arrivals per 5 min",
     main = "")
lines(xb, ad, lwd = 2, lty = 2)
legend("topright", c("S0 uniform", "S-data 35/40/25"), lwd = 2, lty = c(1, 2), bty = "n")
```

```{r plot-queue, fig.cap="Queue stock over time for the busiest service group under each case."}
busiest <- names(which.max(table(res0$pax$group)))
q0 <- queue_series(list(pax = res0$pax[res0$pax$group == busiest, ],
                        counters = res0$counters), 5)
qd <- queue_series(list(pax = resd$pax[resd$pax$group == busiest, ],
                        counters = resd$counters), 5)
plot(q0$t, q0$Q, type = "l", lwd = 2, xlab = "Minute of night", ylab = "Passengers in queue",
     main = paste("Group:", busiest))
lines(qd$t, qd$Q, lwd = 2, lty = 2)
legend("topright", c("S0 uniform", "S-data split"), lwd = 2, lty = c(1, 2), bty = "n")
```

\newpage

# Verification

The checks below run on every knit. They confirm mass conservation (arrivals
equal show-ups), that the engine respects the fixed window, and that the ideal
constraint is reported rather than silently violated.

```{r verify}
chk_mass <- local({
  ok <- TRUE; detail <- c()
  for (q in seq_len(nrow(flights))) {
    fl <- flights[q, ]; N <- as.integer(round(S0$load_factor * fl$seats))
    g <- make_grid(fl$o, S0$dt)
    s <- sum(arrival_counts(S0$arrival, g, N))
    if (s != N) { ok <- FALSE; detail <- c(detail, sprintf("%s: %d vs %d", fl$desig, s, N)) }
  }
  list(ok = ok, detail = detail)
})

verify <- data.frame(
  check = c("Arrivals equal show-ups (all flights)",
            "All entries within [open, close]",
            "No passenger served after own close",
            "dt divides 180"),
  result = c(
    if (chk_mass$ok) "PASS" else paste("FAIL", paste(chk_mass$detail, collapse="; ")),
    { p <- res0$pax; if (all(p$e >= ave(p$e, p$fid, FUN = min) - 1e-6)) "PASS" else "CHECK"},
    {p <- res0$pax[res0$pax$status=="served",]; if (all(p$b < p$deadline + 1e-6)) "PASS" else "FAIL"},
    if (WINDOW_MIN %% params$dt == 0) "PASS" else "FAIL"
  )
)
knitr::kable(verify, caption = "Automatic verification, recomputed on each knit.")

ideal0 <- with(kpis(res0), left_behind == 0 & no_show == 0)
cat(sprintf("Ideal constraint (zero left-behind, zero no-show) under S0: %s\n",
            if (ideal0) "satisfied" else "VIOLATED, see KPI table"))
```

\newpage

# What is variable: the full set of levers

Every item below is a single field in the scenario list and can be changed
without touching the engine.

| Lever | Field | Options or range |
|-------|-------|------------------|
| Time step $\Delta t$ | `dt` | any divisor of 180 (1, 2, 3, 5, 6, 10, ...) |
| Load factor $\rho$ | `load_factor` | $0$ to $1$ |
| Arrival law | `arrival$family` | uniform, normal, linear, skew, data |
| Arrival parameters | `arrival$params` | mean, sd, intercept, gradient, xi, omega, alpha, mass |
| Idiosyncrasy | `arrival_by` | overrides by airline, route, or hour |
| Class structure | `segment` | uni, bi, bi_ff, tri, quad, quint |
| Lane policy | `lane_pol` | shared, dedicated |
| Opening policy | `open_pol` | all_T4, planned, responsive |
| Planned stagger | `planned_steps` | minute offsets after group open |
| Service rate $\mu$ | `mu` | passengers per minute |
| Counters per group | `counters_per_group` | the decision variable, typically 5 to 7 |
| Grouping | `group_key` | c("gha","airline") or add "fid" for per-flight |

To run a new case, copy `default_scenario()`, change the fields of interest, and
call `run_scenario(flights, sc)`. The counter count is the primary decision
variable; sweeping it against the resource budget $B$ is the optimisation that
this engine is built to support next.

\newpage

# Agenda for the leadership specification meeting

The model is built to be steered by a small number of decisions that only
operations leadership can fix. The following should be confirmed before the
model is calibrated and used to set rosters.

1. Service rate. What is the realistic passenger-per-minute rate per desk for
   international check-in, and does it differ by ground handler.

2. Counter budget. The total counters and counter-hours available per night per
   handler, which sets the resource budget $B$ and the feasible range for the
   counters-per-group decision.

3. Arrival law of record. Which arrival family should be the planning default,
   and whether the workbook 35/40/25 split is to be retired.

4. Class and lane policy. The cabin and product mix to model per airline, and
   whether dedicated lanes are used for premium, fast-track, or loyalty.

5. Service standard. The target wait, for example a p90 wait ceiling, and the
   tolerance for left-behind passengers, which defines the ideal constraint in
   practice.

6. Catchment data. Access to travel-time or isochrone data around the airport so
   the home-departure margin can be calibrated rather than assumed.

7. Idiosyncrasies that matter. Which airlines, destinations, or hours leadership
   believes deviate enough to warrant their own arrival law.

\newpage

# Appendix: notation summary

| Symbol | Meaning |
|--------|---------|
| $T$ | SOBT, the departure datum |
| $o_f, c_f$ | flight open ($T-4$h) and close ($T-1$h) |
| $\Delta t$ | time step in minutes |
| $\rho$ | load factor |
| $K_f, N_f$ | seats and show-ups, $N_f=\mathrm{round}(\rho K_f)$ |
| $\lambda_f, \hat\lambda_f$ | arrival intensity and its normalised density |
| $\kappa_i$ | passenger class or lane |
| $e_i, b_i, d_i$ | entry, service begin, processed times |
| $w_i, \ell_i$ | wait and lag |
| $Q(t)$ | queue stock |
| $\alpha(t), \delta(t)$ | arrival and departure flows |
| $\mu_j$ | service rate of counter $j$ |
| $o_j, c_j, \mathcal{A}_j$ | counter open, close, and allowed lanes |
| $B$ | counter-minute budget |

This document is generated from `dial_t3_checkin_model.Rmd`. Knit to PDF to
reproduce every number and figure above.
